### 2019-12-28 interim/sections/background/session_types.tex
Interim write-up: multiparty session types

$$
\infer[\mbox{\rulename{Comm}}]{\piout{a}{v} \mid \piin{a}{x}.P ~\longrightarrow ~ P[v/x] }{}
$$

@@

\subsection{Binary Session Types}
% Introduce synchronous pi calculus with branching and selection
A \term{binary session} is a parallel composition of two processes, each representing a distinct participant.
In the context of web applications, a binary session would describe the interactions between client and server.
Without loss of generality, a \term{session} represents the sequence of send and receive actions of a single participant.

@@

$$
\infer[\mbox{\rulename{Comm}}]{
	\parti{p} :: \sel{\parti{q}}{l}{e_1 \dots e_n}.~P \mid 
	\parti{q} :: \bra{\parti{p}}{l_i(x_1 \dots x_n):Q_i}_{i\in I} ~\longrightarrow~ \parti{p} :: P \mid \parti{q} ::Q_j[v_k/x_k]^n_{k=1}
}{
	\begin{array}{cccccc}
	\exists	j \in I. l_j = l
	&
	e_1 \downarrow v_1 
	& \dots 
	& e_n \downarrow v_n
	&
	& \parti{p} \neq \parti{q}
	\end{array}
}
$$

@@

$$
\infer[\mbox{\rulename{Ty-Bra}}]{
	\Gamma \vdash \bra{\parti{p}}{l_i(x_1 \dots x_n):P_i}_{i\in I} : \tybra{\parti{p}}{l_i(U_1 \dots U_n):S_i}_{i\in I}
	}{
	\begin{array}{ccc}
	\forall i \in I.
	&
	&
    \Gamma, x_1: U_1, \dots, x_n: U_n \vdash P_i : S_i		
	\end{array}
}
$$

$$
\infer[\mbox{\rulename{Ty-Sel}}]{
	\Gamma \vdash \sel{\parti{p}}{l}{e_1 \dots e_n}.P : \tysel{\parti{p}}{l(U_1 \dots U_n): S}
	}{
	\begin{array}{cccc}
	\Gamma \vdash e_1 : U_1
	&
	\dots
	&
	\Gamma \vdash e_n : U_n
	&
    \Gamma \vdash P : S
	\end{array}
}
$$

@@

$$
\infer[\mbox{\rulename{MTy}}]{
	\vdash \parti{p} :: P \mid 
	\parti{q} :: Q
}{
	\begin{array}{lr}
	\cdot \vdash P: S
	&
	\cdot \vdash Q: \overline{S}
	\end{array}
}
$$

@@

$$
\begin{array}{rl}
S_{\text{Client}} &= \tysel{\parti{Server}}{Succ(\texttt{int}): \tybra{\parti{Server}}{Res(\texttt{int}): \mathbf{end}}} \\
S_{\text{Server}} &= \tybra{\parti{Client}}{Succ(\texttt{int}): \tysel{\parti{Client}}{Res(\texttt{int}): \mathbf{end}},~Quit(): \mathbf{end}}
\end{array}
$$

@@

$$
\infer[\mbox{\rulename{Ty-Sub}}]{
	\Gamma \vdash P : S'
}{
	\begin{array}{lr}
	\Gamma \vdash P: S
	&
	S \leqslant S'
	\end{array}
}
$$

@@

\subsection{Multiparty Session Types}
\begin{itemize}
\item Motivate the extension from BST to MPST (example of battleship game, existing distributed protocols that involve multiple participants?)
\item Introduce global type, projection
\item How does the notion of compatibility (which used to be duality) adapt to MPST?
\end{itemize}

G ::= & & \text{Global Types} \\
     & \mathbf{end} & \text{Termination} \\
\mid & \parti{p} \to \parti{q}:[U]; G & \text{Termination} \\
\mid & \tysend{\parti{p}}{U};~S & \text{Value Send} \\
\mid & \tyrecv{\parti{p}}{U};~S & \text{Value Receive} \\
\mid & \tybra{\parti{p}}{l_i:S_i}_{i\in I} & \text{Branching} \\
\mid & \tysel{\parti{p}}{l_i:S_i}_{i\in I} & \text{Selection} \\
\mid & \mu \mathbf{t}.~S & \text{Recursive Type} \\
\mid & \mathbf{t} & \text{Type Variable} \\
\end{array}
\]

@@

$$
\infer[\mbox{\rulename{MTy}}]{
	\vdash \prod_{i \in I} \parti{p_i} :: P_i ~: G
}{
	\begin{array}{llll}
	\forall i \in I. 
	&
	\cdot \vdash P_i: G \upharpoonright \parti{p_i}
	&
	&
	\text{pt}(G) \subseteq \left\{\parti{p_i} \mid i \in I \right\}
	\end{array}
}
$$

>>>>

\[
\infer[\mbox{\rulename{Comm}}]{\piout{a}{v} \mid \piin{a}{x}.P ~\longrightarrow ~ P[v/x] }{}
\]

@@

\subsection{Binary Session Types}\label{section_bst}
% Introduce synchronous pi calculus with branching and selection
A \term{binary session} is a parallel composition of two processes, each representing a distinct participant.
In the context of web applications, a binary session would describe the interactions between client and server.
Without loss of generality, a \term{session} represents the sequence of send and receive actions of a single participant.

@@

\[
\infer[\mbox{\rulename{Comm}}]{
	\parti{p} :: \sel{\parti{q}}{l}{e_1 \dots e_n}.~P \mid 
	\parti{q} :: \bra{\parti{p}}{l_i(x_1 \dots x_n):Q_i}_{i\in I} ~\longrightarrow~ \parti{p} :: P \mid \parti{q} ::Q_j[v_k/x_k]^n_{k=1}
}{
	\begin{array}{cccccc}
	\exists	j \in I. l_j = l
	&
	e_1 \downarrow v_1 
	& \dots 
	& e_n \downarrow v_n
	&
	& \parti{p} \neq \parti{q}
	\end{array}
}
\]

@@

\[
\infer[\mbox{\rulename{Ty-Bra}}]{
	\Gamma \vdash \bra{\parti{p}}{l_i(x_1 \dots x_n):P_i}_{i\in I} : \tybra{\parti{p}}{l_i(U_1 \dots U_n):S_i}_{i\in I}
	}{
	\begin{array}{ccc}
	\forall i \in I.
	&
	&
    \Gamma, x_1: U_1, \dots, x_n: U_n \vdash P_i : S_i		
	\end{array}
}
\]

\[
\infer[\mbox{\rulename{Ty-Sel}}]{
	\Gamma \vdash \sel{\parti{p}}{l}{e_1 \dots e_n}.P : \tysel{\parti{p}}{l(U_1 \dots U_n): S}
	}{
	\begin{array}{cccc}
	\Gamma \vdash e_1 : U_1
	&
	\dots
	&
	\Gamma \vdash e_n : U_n
	&
    \Gamma \vdash P : S
	\end{array}
}
\]

@@

\[
\infer[\mbox{\rulename{MTy}}]{
	\vdash \parti{p} :: P \mid 
	\parti{q} :: Q
}{
	\begin{array}{lr}
	\cdot \vdash P: S
	&
	\cdot \vdash Q: \overline{S}
	\end{array}
}
\]

@@

\[
\begin{array}{rl}
S_{\text{Client}} &= \tysel{\parti{Server}}{\text{Succ}(\texttt{int}): \tybra{\parti{Server}}{\text{Res}(\texttt{int}): \mathbf{end}}} \\
S_{\text{Server}} &= \tybra{\parti{Client}}{\text{Succ}(\texttt{int}): \tysel{\parti{Client}}{\text{Res}(\texttt{int}): \mathbf{end}},~\text{Quit}(): \mathbf{end}}
\end{array}
\]

@@

\[
\infer[\mbox{\rulename{Ty-Sub}}]{
	\Gamma \vdash P : S'
}{
	\begin{array}{lr}
	\Gamma \vdash P: S
	&
	S \leqslant S'
	\end{array}
}
\]

@@

\subsection{Multiparty Session Types}

% Motivate the extension from BST to MPST (example of battleship game, existing distributed protocols that involve multiple participants?)
Whilst binary session types provide communication guarantees between exactly 2 participants, distributed systems generally involve more parties in practice. This is equally relevant in interactive web applications, as motivated by the \textit{Battleships} game example in \cite{PureScript2019} where the server coordinates interactions between two players.

Whilst there is a natural syntactical extension to our session calculus for describing multiparty sessions\footnote{We also adopt the shorthand $\mathcal{M} ::= \prod^n_{i=1} \parti{p_i} :: P_i$ as used in the literature.} as

\[
\begin{array}{rl}
\mathcal{M} ::= & \parti{p_1} :: P_1~\mid~\parti{p_2} :: P_2~\mid~\dots~\mid \parti{p_n} :: P_n
\end{array}
\]

the same cannot be said for the binary session typing discipline, particularly with respect to duality. The same notion of duality does not extend to the decomposition of multiparty interactions into multiple binary sessions:  \cite{C406Lecture} and \cite{MPST} both present counterexamples of well-typed binary sessions that, when composed to represent a multiparty session, results in communication errors thus violating guarantees of well-typed sessions.

Honda et al. presents \term{multiparty session types} in \cite{MPAST} to extend the binary session typing discipline for sessions involving more than 2 participants, whilst redefining the notion of compatibility in this multiparty context. Multiparty session types are defined in terms a \term{global type}, which provides a bird's eye view of the communication protocol describing the interactions between pairs of participants. Figure \ref{fig:mpst} defines the syntax of global types inspired by \cite{MPST} and adapted to be compatible with our session calculus presented in \mathref{\ref{section_bst}}.

% Introduce global type, projection
\begin{figure}[!hb]
\doublespacing
\[
\begin{array}{rlr}

G ::= & & \text{Global Types} \\
     & \mathbf{end} & \text{Termination} \\
\mid & \parti{p} \to \parti{q}:\left\{l_i(U_1 \dots U_n).~G_i\right\}_{i \in I} & \text{Message Exchange} \\
\mid & \mu \mathbf{t}.~G & \text{Recursive Type} \\
\mid & \mathbf{t} & \text{Type Variable} \\
\end{array}
\]

@@

To check the conformance of a participant's process against the protocol specification described by the global type, we \term{project} the global type onto the participant to obtain a session type that only preserves the interactions described by the global type that pertain to the participant. Projection is defined by the $\upharpoonright$ operator, more commonly seen in literature in its infix form as $G \upharpoonright \parti{p}$ describing the projection of global type $G$ for participant $\parti{p}$. Intuitively, the projected local type of a participant describes the protocol from the viewpoint of the participant.

More formally, projection can be interpreted as a \textit{partial function} ${\upharpoonright} :: G \times \parti{p} \rightharpoonup S$, as the projection for a participant may be undefined for an ill-formed global type; \cite{C406Lecture} presents examples of where this is the case, and \cite{MPST} presents the formal definition of projection.

% How does the notion of compatibility (which used to be duality) adapt to MPST?
The notion of compatibility in multiparty session types is still captured by \rulename{MTy}, but adapted to consider the local projections for all participants as supposed to dual types in the binary case.

\[
\infer[\mbox{\rulename{MTy}}]{
	{\displaystyle \vdash \prod^{}_{i \in I} \parti{p_i} :: P_i ~: G} 
}{
	\begin{array}{llll}
	\forall i \in I. 
	&
	\cdot \vdash P_i: G \upharpoonright \parti{p_i}
	&
	&
	\text{pt}(G) \subseteq \left\{\parti{p_i} \mid i \in I \right\}
	\end{array}
}
\]

For a multiparty session $\mathcal{M} = \prod_{i \in I}\parti{p_i} :: P_i$ to be well-typed by a global type $G$, we require:

\begin{enumerate}
\item All participant processes $\parti{p_i} :: P_i$ to be well-typed with respect to their corresponding well-defined projection $G \upharpoonright \parti{p_i}$, and 
\item $G$ does not describe interactions with participants not defined in $\mathcal{M}$ 
\end{enumerate}

Well-typed multiparty session enjoys the following communication guarantees as outlined in \cite{GentleMPST}:

\begin{itemize}
\item \textbf{Communication safety}: The types of sent and expected messages will always match.
\item \textbf{Protocol fidelity}: The exchange of messages between processes will abide by the global protocol.
\item \textbf{Progress}: Messages sent by a process will be eventually received, and a process waiting for a message will eventually receive one; this also means there will not be any sent but unreceived messages.
\end{itemize}

This motivates an elegant, decentralised solution for checking protocol conformance in practice: once the global type for the protocol is defined, local processes can verify their implementation against their corresponding projection in isolation, independent of each other. Figure \ref{fig:mpst_workflow}\todo{Create figure} illustrates this diagrammatically.

\begin{figure}[!h]
\caption{Type checking with Multiparty Session Types}
\label{fig:mpst_workflow}
\end{figure}
